---
title: "Simulating Data with Causal Graphs"
eval: false
---

```{r 03_setup, eval=TRUE, message=FALSE}
library(tidyverse)
library(scales)
library(broom)
library(dagitty)
```

# Goals {.unnumbered}

- Explain how associations arise from different structures in causal graphs
- Gain confidence exploring statistical properties via simulation

<br>

**You can download a template file for this activity [here](templates/03-causal-graphs-simulation.qmd).**

Slides for today are available [here](https://docs.google.com/presentation/d/1dmulGgy3gV0NwJEjcOS9QpliEmfiJfgFvYwPShlq_DE/edit?usp=sharing).

# Small Group Discussion {.unnumbered}

## Prompt 1 {.unnumbered}

a. Does the following causal diagram reflect a causal relationship between treatment A and outcome Y? Will A and Y likely appear related (through modeling results or exploratory plots)?

```{r 03_p1_dag1, eval=TRUE, echo=FALSE, fig.width=5, fig.height=3, fig.align="center"}
dag <- dagitty("
dag {
bb=\"0,0,1,1\"
A [exposure,pos=\"0.250,0.500\"]
C [pos=\"0.500,0.250\"]
Y [outcome,pos=\"0.750,0.500\"]
C -> A
C -> Y
}
")

plot(dag)
```

b. Repeat part a for the causal diagram below. Do you think the observed relationship between A and Y will be different from what you might have observed if the previous causal diagram were the true data generating process?

```{r 03_p1_dag2, eval=TRUE, echo=FALSE, fig.width=5, fig.height=3, fig.align="center"}
dag <- dagitty("
dag {
bb=\"0,0,1,1\"
A [exposure,pos=\"0.250,0.500\"]
C [pos=\"0.500,0.250\"]
Y [outcome,pos=\"0.750,0.500\"]
A -> Y
C -> A
C -> Y
}
")

plot(dag)
```

## Prompt 2 {.unnumbered}

The following causal diagram has a fork (`A <- C -> Y`), chain (`A -> M -> Y`), and collider (`A -> S <- Y`).

`S` is a binary variable that equals 1 if the data for a unit of observation are made available to us and equals 0 otherwise. Thus, `S = 1` for the data we have access to.

Considering the fork, chain, and collider separately:

- Will treatment A and outcome Y appear related?
- If they will appear related, is there a way to make them unrelated?

If we could isolate what is happening in just one of the parts, would you choose the fork, chain, or collider?

```{r 03_p2_dag, eval=TRUE, echo=FALSE, fig.width=5, fig.height=3, fig.align="center"}
dag <- dagitty("
dag {
bb=\"0,0,1,1\"
A [exposure,pos=\"0.250,0.500\"]
C [pos=\"0.500,0.250\"]
M [pos=\"0.500,0.500\"]
S [selected,pos=\"0.500,0.750\"]
Y [outcome,pos=\"0.750,0.500\"]
A -> M
A -> S
C -> A
C -> Y
M -> Y
Y -> S
}
")

plot(dag)
```

## Prompt 3 {.unnumbered}

For the causal diagram below, suppose that you wanted to generate (simulate) data according to this data generating process. In what order would you create the variables?

```{r 03_p3_dag, eval=TRUE, echo=FALSE, fig.width=5, fig.height=3, fig.align="center"}
dag <- dagitty("
dag {
bb=\"0,0,1,1\"
A [exposure,pos=\"0.250,0.500\"]
C [pos=\"0.500,0.250\"]
Y [outcome,pos=\"0.750,0.500\"]
Z [pos=\"0.613,0.375\"]
A -> Y
C -> A
C -> Z
Z -> Y
}
")

plot(dag)
```

## Prompt 4 {.unnumbered}

Suppose that a quantitative variable C and a categorical variable Z were both causes of a binary outcome Y. How could you use a logistic regression model formula to generate Y?


<br>


# Notes {.unnumbered}

## Structure of causal graphs {.unnumbered}

The fact that causal graphs depict a data-generating process means that it conveys statistical information about the relationships between variables.

3 structures form the basis of every pattern in a causal graph:

1. Chains: `A -> B -> C`
2. Forks: `A <- B -> C` (B is a common cause)
3. Colliders: `A -> B <- C` (B is a common effect)


<br><br><br><br>


## Examples {.unnumbered}

Note: I asked ChatGPT to help me generate examples of these structures in different contexts with the following prompt:

> Give me an example of a [chain,fork,collider] causal graph structure in the following contexts: mental health, sociology, biology, medicine, public health, economics, politics.

**Chains:**

- Mental Health: Childhood Trauma -> Depression -> Substance Abuse
- Sociology: Education Level -> Employment Status -> Income Level
- Biology: Gene Mutation -> Protein Dysfunction -> Disease Onset
- Medicine: Infection -> Immune Response -> Fever
- Public Health: Air Pollution -> Respiratory Problems -> Hospital Admissions
- Economics: Interest Rate Changes -> Investment Levels -> Economic Growth
- Politics: Political Campaign -> Public Opinion -> Election Outcome

**Forks:**

- Mental Health: Depression <- Genetic Predisposition -> Anxiety
- Sociology: Children's Socioeconomic Status <- Parental Education Level -> Children's Educational Attainment
- Biology: Weight Gain <- Hormonal Imbalance -> Fatigue
- Medicine: Cardiovascular Disease <- Poor Diet -> Obesity
- Public Health: Health Outcomes <- Socioeconomic Status -> Access to Healthcare
- Economics: Consumer Spending <- Economic Recession -> Unemployment Rates
- Politics: Voting Behavior <- Political Polarization -> Media Consumption

**Colliders:**

- Mental Health: Genetic Predisposition -> Depression <- Life Stressors
- Sociology: Educational Attainment -> Income Level <- Social Network
- Biology: Gene Mutation -> Disease Onset <- Environmental Factors
- Medicine: Lack of Physical Activity -> Obesity <- Poor Diet
- Public Health: Air Pollution -> Respiratory Problems <- Smoking
- Economics: Economic Policies -> Inflation Rate <- Global Market Conditions
- Politics: Campaign Funding -> Election Outcome <- Public Opinion


<br><br><br><br>


## Marginal and conditional (in)dependence in causal graphs {.unnumbered}

- Chains, forks, and colliders have different patterns of marginal and conditional independence and dependence.
- These patterns are the core of what makes something an alternate explanation and how we can address that alternate explanation through analysis (coming next time).
- The marginal/conditional (in)dependence properties are not intuitive, but we can see them play out via **simulation**.


<br><br><br><br>


## Simulation {.unnumbered}

- When conducting simulations, we are in control of the data-generating process.
- We control the distribution of all variables and how they are related to each other.
- We can simulate data from a causal graph by:
    1. Simulate variables that don't have causes (called **exogenous variables**)
        - Example: In the chain `X -> Y -> Z`, we would simulate `X` first.
    2. Simulate the variables caused by the exogenous variables. Then simulate the variables caused by these variables, etc.
        - Example: In the chain `X -> Y -> Z`, after simulating `X`, simulate `Y` then `Z`.


<br><br><br><br>


## Simulating quantitative and categorical variables {.unnumbered}

[This table](https://evalsp24.classes.andrewheiss.com/example/random-numbers.html#summary) at the end of our reading provides a nice summary of simulation functions.

For quantitative variables we can use:

- `rnorm()`
- `runif()`
- `rescale()` (This function is part of the `scales` package.)

For categorical variables we can use:

- `rbinom()`: For binary variables
- `sample()`: For 3+ categories

We can look up the documentation for functions by entering `?function_name` in the Console. Use the documentation for these functions to complete the following exercises:

```{r}
# Set seed for random number generator for reproducible results
set.seed(451)

# Simulate 1000 normally distributed numbers with mean 10 and standard deviation 2. Store as variable X
X <- 

# Check distribution of X with a plot (code is complete)
ggplot(data.frame(X), aes(x = X)) +
    geom_histogram()

# Simulate 1000 uniformly distributed numbers between 10 and 20. Store as variable Y
Y <- 

# Check distribution of Y with a plot (code is complete)
ggplot(data.frame(Y), aes(x = Y)) +
    geom_histogram()


# Rescale X to be between 0 and 1. Store as variable Z
Z <- 

# Check distribution of Y with a plot (code is complete)
ggplot(data.frame(Z), aes(x = Z)) +
    geom_histogram()
```


<br><br><br><br>


## Simulating dependence {.unnumbered}

We can use regression model formulas to have the mean of a variable depend on its causes.

For the mean of quantitative variables:

$$
E[\text{score} \mid \text{hoursPractice}, \text{reviewSession}] = 40 + 3\times\text{hoursPractice} + 10\times \text{reviewSessionYes}
$$
The mean of a binary (0/1) variable is the probability that it equals 1. We can use a similar regression model formula to determine this probability and use `rescale()` to force this number to be between 0 and 1.


<br><br><br><br>


## Example {.unnumbered}

We'll simulate data from this causal graph:

```{r 03_example_dag, eval=TRUE}
dag <- dagitty("dag {
bb=\"0,0,1,1\"
3MonthStress [outcome,pos=\"0.800,0.500\"]
MindfulBreathing [exposure,pos=\"0.200,0.500\"]
PriorStress [pos=\"0.400,0.250\"]
Yoga [pos=\"0.600,0.250\"]
MindfulBreathing -> 3MonthStress
PriorStress -> 3MonthStress
PriorStress -> MindfulBreathing
Yoga -> 3MonthStress
Yoga -> MindfulBreathing
}")
plot(dag)
```

Simulation code:

```{r eval=TRUE}
set.seed(451)
n <- 1000
sim_data <- tibble(
    prior_stress = runif(n, min = 0, max = 10),
    yoga = rbinom(n, size = 1, prob = 0.15),
    p_mindful_breathing = rescale(1 + 3*prior_stress + 25*yoga, to = c(0,1)),
    mindful_breathing = rbinom(n, size = 1, prob = p_mindful_breathing),
    mean_stress_3_month = prior_stress - 3*yoga - 5*mindful_breathing,
    noise_stress_3_month = rnorm(n, mean = 0, sd = 2),
    stress_3_month = mean_stress_3_month + noise_stress_3_month
)
```

It's important to use plots and models to check the relationships in the data we generated:

```{r 03_example_checks, eval=TRUE}
ggplot(sim_data, aes(y = factor(mindful_breathing), x = prior_stress)) +
    geom_boxplot()

ggplot(sim_data, aes(fill = factor(mindful_breathing), x = factor(yoga))) +
    geom_bar(position = "fill")

mod <- lm(stress_3_month ~ yoga + prior_stress + mindful_breathing, data = sim_data)
summary(mod)
```


<br><br><br><br>


## Long-run properties via simulation {.unnumbered}

Suppose that X and Y are independent quantitative variables. Suppose that we fit the model below:

$$E[Y \mid X] = \beta_0 + \beta_1 X$$

- Let our significance level (type I error rate) be 0.05. What would you expect about the p-value for the coefficient on X?
- Now suppose that you have 1000 different samples of X and Y and fit the above model on each sample. What would you expect about the 1000 p-values?

::: {.callout-note appearance="simple" collapse="true" title="Response"}
If X and Y are independent, the null hypothesis $H_0: \beta_1 = 0$ is true. A significance level (type I error rate) of 0.05 means that we're choosing our test statistic threshold to be high enough that, if the null hypothesis were true, we would (incorrectly) reject the null 5% of the time.

We reject the null when p-value < 0.05. Since the null is true in this case, this will happen (p-value < 0.05) in 5% of samples. So we're expecting 50 of the 1000 samples to give p-values less than 0.05.
:::



<br><br><br><br>



# Exercises {.unnumbered}

Throughout these exercises, use a sample size of 10,000, a significance level of 0.05, and 1000 simulations.

Use linear regression model formulas to create relationships between variables. Don't use nonlinearity or interaction terms. (We're using this restriction today because we'll be using linear regression models to test for conditional independence and dependence.) The specific numbers in your model formulas are up to you, but don't worry too much about ensuring that the distributions and relationships are fully realistic.

## Exercise 1: chains {.unnumbered}

### Part a: Simulate {.unnumbered}

In the code chunk below, write code to simulate a single dataset from the following chain:

Age -> Total hours reading practice -> Reading ability

```{r}
# Simulate data from the chain
# Store data in a dataset called sim_data_chain
# Call your variables age, practice_hrs, read_ability


# Use plots to verify that univariate distributions look ok and that the direct causal relationships look right

```

Copy and paste the code that creates the `sim_data_chain` dataset into the space indicated in the function below. Then complete the next block of code to fit an appropriate model for checking how age and reading ability are related conditional on hours of practice. The remainder of the code looks at the p-values from 1000 simulations.

```{r}
simulate_chain <- function() {
    n <- 10000
    # Paste your code for creating your simulated chain data here
    sim_data_chain <- ___
    
    # Fit a linear regression model that allows you to see 
    # the relationship between read_ability and age 
    # conditional on practice_hrs
    mod <- ___
    
    # Pull out the p-value for the age coefficient
    tidy(mod) %>% 
        filter(term=="age") %>% 
        pull(p.value)
}

# Repeat the simulation 1000 times (generate 1000 datasets)
set.seed(451)
system.time({
chain_pvals <- replicate(1000, {
    simulate_chain()
})
})

# How many times were the p-values less than 0.05?
# Is this what you expected?
table(chain_pvals < 0.05)
```

### Part b: Explain {.unnumbered}

Your target audience is a STAT 155 student. Provide a non-technical explanation for the relationship seen between age and reading ability after adjusting for/conditioning on hours of reading practice.


<br><br>


## Exercise 2: forks {.unnumbered}

### Part a: Simulate {.unnumbered}

In the code chunk below, write code to simulate a single dataset from the following fork:

% of population wearing shorts <- Temperature -> Ice cream sales

```{r}
# Simulate data from the fork
# Store data in a dataset called sim_data_fork
# Call your variables perc_shorts, temperature, ice_cream_sales


# Use plots to verify that univariate distributions look ok and that the direct causal relationships look right

```

Copy and paste the code that creates the `sim_data_fork` dataset into the space indicated in the function below. Then complete the next block of code to fit an appropriate model for checking how age and reading ability are related conditional on hours of practice. The remainder of the code looks at the p-values from 1000 simulations.

```{r}
simulate_fork <- function() {
    n <- 10000
    # Paste your code for creating your simulated fork data here
    sim_data_fork <- ___
    
    # Fit a linear regression model that allows you to see 
    # the relationship between % shorts and ice cream sales
    # conditional on temperature
    mod <- ___
    
    # Pull out the p-value for the perc_shorts coefficient
    tidy(mod) %>% 
        filter(term=="perc_shorts") %>% 
        pull(p.value)
}

# Repeat the simulation 1000 times (generate 1000 datasets)
set.seed(451)
system.time({
fork_pvals <- replicate(1000, {
    simulate_fork()
})
})

# How many times were the p-values less than 0.05?
# Is this what you expected?
table(fork_pvals < 0.05)
```

### Part b: Explain {.unnumbered}

Your target audience is a STAT 155 student. Provide a non-technical explanation for the relationship seen between shorts prevalence and ice cream sales after adjusting for/conditioning on temperature.


<br><br>


## Exercise 3: colliders {.unnumbered}

### Part a: Simulate {.unnumbered}

In the code chunk below, write code to simulate a single dataset from the following collider:

Smoking -> Hospitalization <- COVID-19 Infection

```{r}
# Simulate data from the collider
# Store data in a dataset called sim_data_collider
# Call your variables covid, hosp, smoking


# Use plots to verify that univariate distributions look ok and that the direct causal relationships look right
# Also use a plot and/or model to verify the marginal relationship between covid and smoking

```

Copy and paste the code that creates the `sim_data_collider` dataset into the space indicated in the function below. Then complete the next block of code to fit an appropriate model for checking how age and reading ability are related conditional on hours of practice. The remainder of the code looks at the p-values from 1000 simulations.

- Note: You may have to increase the magnitude of the `smoking -> hosp` effect and the magnitude of the `covid -> hosp` effect to see the expected behavior.
- If you end up updating the numbers in your model formulas in light of the above note, try adding **arguments** to your function like this:
    - `simulate_collider <- function(smoking_effect, covid_effect)`
    - This allows `smoking_effect` and `covid_effect` to be inputs to your function. Replace the appropriate numbers in your `sim_data_collider` code to use `smoking_effect` and `covid_effect`.
    - You'll then need to update the 1000 simulations part to something like:

```r
set.seed(451)
system.time({
collider_pvals <- replicate(1000, {
    simulate_collider(smoking_effect = 10, covid_effect = 10)
})
})
```

```{r}
simulate_collider <- function() {
    n <- 10000
    # Paste your code for creating your simulated collider data here
    sim_data_collider <- ___
    
    # Fit a linear regression model that allows you to see 
    # the relationship between smoking and covid
    # conditional on hospitalization
    mod <- ___
    
    # Pull out the p-value for the smoking coefficient
    tidy(mod) %>% 
        filter(term=="smoking") %>% 
        pull(p.value)
}

# Repeat the simulation 1000 times (generate 1000 datasets)
set.seed(451)
system.time({
collider_pvals <- replicate(1000, {
    simulate_collider()
})
})

# How many times were the p-values less than 0.05?
# Is this what you expected?
table(collider_pvals < 0.05)
```

### Part b: Explain {.unnumbered}

Your target audience is a STAT 155 student. Provide a non-technical explanation for the relationship seen between smoking and COVID infection after adjusting for/conditioning on hospitalization.

(BTW: The motivation for this exercise came from considerable discussion in the causal inference community about risk factors studies early in the COVID pandemic. See [Tweet 1](https://x.com/jonathanasterne/status/1268117716713512960), [Tweet 2](https://x.com/EpiEllie/status/1258607277357006849), and [Tweet 3](https://x.com/David_Simons_UK/status/1258543096461004801) to learn more.)



<br><br>


## Exercise 4: beyond building blocks {.unnumbered}

Can we extend building block thinking to longer, more complex structures? Let's investigate here (conceptually, no simulation).

a. Consider the longer structure `A <- B <- C -> D`. What do you expect about marginal/conditional (in)dependence of `A` and `D`? Explain.

b. Consider the longer structure `A -> B <- C <- D -> E`. What do you expect about marginal/conditional (in)dependence of `A` and `E`? Explain.


<br><br><br><br>


# Reflection {.unnumbered}

- How do the ideas from today fit into the big picture so far?
- What questions remain?

