---
title: "Weighting"
author: "YOUR NAME"
format: 
  html:
    embed-resources: true
    toc: true
---

# Exercises

We will continue looking at the data from the National Supported Work Demonstration project. What was the effect of this job training program on income following the program?

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(WeightIt)
library(MatchIt)
library(cobalt)
library(marginaleffects)

data(lalonde)
```

**Variables of interest**:

- Treatment/exposure: `treat` (Individual was assigned to the job training program, 1 = yes, 0 = no)
- Outcome: `re78` (Individual's income in 1978, in US dollars)

**Covariates**: all are variables that happen prior to treatment (also called **pre-treatment covariates**)

- `age`: age in years
- `educ`: education in number of years of schooling
- `race`: the individual's race/ethnicity, (Black, Hispanic, or White)
- `married`: an indicator for marital status (1 = married, 0 = not married)
- `nodegree`: an indicator for whether the individual has a high school degree (1 = no degree, 0 = degree)
- `re74`: income in 1974, in US dollars
- `re75`: income in 1975, in US dollars





## Exercise 1

First, let's see how we would compute weights by hand.

The code below fits a logistic regression model to estimate the propensity score for the `lalonde` data:

```{r}
logistic_mod <- glm(treat ~ age + educ + race + married + nodegree + re74 + re75, data = lalonde, family = "binomial")
```

*Food for thought:* Is the above model a *good* model for the propensity score? We can check by assessing balance statistics after weighting. We won't do that here, but we'll explore how this in Exercise 2.

Next we use our model to obtain the propensity scores themselves and subsequently our weights. Look through the code below to see how ATE weights are computed. Make sure you understand the process, and clarify anything if needed.

Add to this code to also compute ATT and ATC (ATU) weights.

```{r}
lalonde_with_wts <- lalonde %>% 
    mutate(
        # Obtain propensity scores from logistic regression model (log odds->odds->p)
        ps = predict(logistic_mod, newdata = lalonde, type = "response"),
        w_ate = case_when(
            treat==1 ~ 1/ps, # When A=1, ATE weight is 1/P(A=1|Z)
            treat==0 ~ 1/(1-ps) # When A=0, ATE weight is 1/P(A=0|Z) = 1/(1-P(A=1|Z))
        )
    )
```





## Exercise 2

The `WeightIt` package implements numerous weighting methods for causal inference. It is a parallel to the `MatchIt` package and maintained by some of the same authors, so the structure of code is similar.

The [WeightIt vignette](https://ngreifer.github.io/WeightIt/articles/WeightIt.html) is a useful reference and is worth reading later (the "Introduction" and "Balancing Weights for a Point Treatment" sections).

Open up the `weightit()` documentation by entering `?weightit` in the Console. Implement inverse probability weighting to estimate the ATE, ATT, and ATC using logistic regression to estimate the propensity score. We will continue using all covariates (`age + educ + race + married + nodegree + re74 + re75`) to block noncausal paths.

```{r}
# Don't include any interactions terms in the formula argument for now (for simplicity)
weight_out_ate <- weightit(
    ???
)

weight_out_att <- weightit(
    ???
)

weight_out_atc <- weightit(
    ???
)
```

Check that the weights that you computed by hand are identical to the ones from `weightit()`:

```{r}
# all.equal() checks if 2 vectors are basically identical (give or take some 
# differences in floating point precision)
# We use unname to get rid of the unit IDs that show up as names for the
# vector elements in our manual calculation
all.equal(unname(weight_out_ate$weights), unname(lalonde_with_wts$w_ate))
all.equal(unname(weight_out_att$weights), unname(lalonde_with_wts$w_att))
all.equal(unname(weight_out_atc$weights), unname(lalonde_with_wts$w_atc))
```

Generally speaking, the variability of the weights is a way to quantify their quality. Highly variable weights will lead to highly variable treatment effect estimates. We can check the variability of the estimated weights with the `summary()` function. As you inspect the output below, focus on connecting what you see with different ways to measure variability:

- The range: Weights above 10 and much less than 1 are somewhat extreme
- The coefficient of variation ("Coef of Var") is the standard deviation divided by the mean. Generally, values greater than 1 indicate large variation (large variation relative to the mean).
- MAD is median absolute deviation. It's like standard deviation but robust to outliers. High values indicate high variability.
- Entropy is a quantity from information theory. It measures the amount of "disorder" in a probability distribution. High values indicate high variability.
- Effective sample sizes are displayed just like from `MatchIt`.

```{r}
summary(weight_out_ate)
summary(weight_out_att)
summary(weight_out_atc)
```

We can also use `bal.tab()` and `love.plot()` as we did in matching to inspect covariate balance after weighting. Inspect this output to refamiliarize yourself with the information it conveys.

```{r}
bal.tab(
    weight_out_ate,
    stats = c("mean.diffs", "variance.ratios"),
    int = TRUE,
    thresholds = c("mean.diffs" = 0.1, "variance.ratios" = 2) # Adding this compares each balance statistic to the threshold for a quicker visual summary of what covariates are balanced or not
)
```

```{r}
love.plot(
    weight_out_ate,
    stats = c("mean.diffs", "variance.ratios"),
    int = TRUE,
    stars = "raw",
    thresholds = c("mean.diffs" = 0.1, "variance.ratios" = 2)
)
```

Note: We can consider trimming extreme weights with `trim()` function, but we will not do so here. (More information on trimming is available [here](https://ngreifer.github.io/WeightIt/reference/trim.html).)

We can estimate the treatment effect in a very similar way to what we did for matching. Take a look at the code below, and draw comparisons to the process we used during matching. Also make sure that you feel comfortable interpreting the results. (Interpretations are the same as in the matching setting.)

```{r}
## ATE
fit_ate <- lm_weightit(
    re78 ~ treat * (age + educ + race + married + nodegree + re74 + re75),
    data = lalonde,
    weightit = weight_out_ate
)

# We leave the vcov argument unspecified to be at its default value (TRUE). This estimates standard errors in an approrpriate way as described here: https://ngreifer.github.io/WeightIt/articles/estimating-effects.html#estimating-standard-errors-and-confidence-intervals
avg_comparisons(
    fit_ate,
    variables = "treat",
    newdata = lalonde
)

## ATT
fit_att <- lm_weightit(
    re78 ~ treat * (age + educ + race + married + nodegree + re74 + re75),
    data = lalonde,
    weightit = weight_out_att
)

avg_comparisons(
    fit_att,
    variables = "treat",
    newdata = filter(lalonde, treat==1)
)

## ATC
fit_atc <- lm_weightit(
    re78 ~ treat * (age + educ + race + married + nodegree + re74 + re75),
    data = lalonde,
    weightit = weight_out_atc
)

avg_comparisons(
    fit_att,
    variables = "treat",
    newdata = filter(lalonde, treat==0)
)
```





## Exercise 3

It turns out that when we performed matching, we were using weighting too! When we used matching, a `weights` column was computed in a way that mirrors what inverse probability weighting with the propensity score is doing. Within a subclass (a matched set of treated and control units), the "propensity score" is the proportion of treated units in that subclass. Then the weights are computed with the same formulas that we derived by hand in exercise 1.

Draw a connection between what we were doing with inverse probability weighting with the tree diagrams to the weighting process described here. Check in with the instructor if you're uncertain about this.

Next, run the code below to run exact matching and view the computed weights. Try to verify by hand how the `weights` were computed for subclasses 1 and 2.

```{r}
match_out_exact <- matchit(
    treat ~ age + educ + race + married + nodegree + re74 + re75,
    data = lalonde,
    method = "exact",
    estimand = "ATT",
    normalize = FALSE # To make hand checking the weights easier
)

match_data_exact <- match.data(match_out_exact)

# Look at the matched data and weights
match_data_exact |> 
    arrange(subclass) |> 
    select(treat, weights, subclass) |> 
    View()
```

Finally, fill in the `???` to compute all the weights manually and compare your manually computed weights to the ones provided by `matchit()`. 

```{r}
match_data_exact <- match_data_exact |> 
    group_by(subclass) |> 
    mutate(
        ps = sum(treat)/length(treat), # PS = fraction of units in this subclass that are treated
        weight_manual = if_else(treat==1, 1, ???)
    )

# They should match!
all.equal(match_data_exact$weights, match_data_exact$weight_manual)
```





## Exercise 4

(This exercise is provided for further exploration but focuses on an additional idea that is not a core learning goal for the course.)

When estimating the ATE, one adjustment we can consider for our weights is **weight stabilization**. By multiplying all the weights for the treated units by the same factor $k_1$ and all the weights for the control units by the same factor $k_2$ we can reduce the spread (range) of the weights, which can help in increasing the precision of our effect estimates (but isn't guaranteed).

(If you want to explore this more: Section 12.3 of our supplemental reference [Causal Inference: What If](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/) discusses stabilized IP weighting.)

The `stabilize` argument can be set to `TRUE` to compute stabilized weights. Compare the range of weights with and without stabilization.

```{r}
weight_out_ate_stabilized <- weightit(
    treat ~ age + educ + race + married + nodegree + re74 + re75,
    data = lalonde,
    estimand = "ATE",
    method = "glm",
    stabilize = TRUE
)

summary(weight_out_ate)
summary(weight_out_ate_stabilized)
```

We can compare the standard errors of our treatment effect estimates with and without stabilization. In this case, stabilized weights didn't reduce the variance of the treatment effect estimate, but it generally is a good idea to explore:

```{r}
# Using unstabilized weights
fit_unstabilized <- lm_weightit(
    re78 ~ treat * (age + educ + race + married + nodegree + re74 + re75),
    data = lalonde,
    weightit = weight_out_ate
)

avg_comparisons(
    fit_unstabilized,
    variables = "treat",
    newdata = lalonde
)

# Using unstabilized weights
fit_stabilized <- lm_weightit(
    re78 ~ treat * (age + educ + race + married + nodegree + re74 + re75),
    data = lalonde,
    weightit = weight_out_ate_stabilized
)

avg_comparisons(
    fit_stabilized,
    variables = "treat",
    newdata = lalonde
)
```





## Exercise 5

(This exercise is provided for further exploration but focuses on an additional idea that is not a core learning goal for the course.)

Entropy balancing is another weighting technique that solves a constrained optimization problem to find weights that *exactly* match **moments** of the covariates and minimize the variance of the weights (variance as quantified by a measure called **entropy**):

- The p-th moment of a probability distribution for a random variable $X$ is $E[X^p]$.
    - The 1st moment is the mean (expected value): $E[X]$
    - The 2nd moment $E[X^2]$ is used in computing the variance ($\text{Var}(X) = E[X^2] - (E[X])^2$)

Below we run entropy balancing to balance twice:

- First time: We balance the first moment (the mean) of each covariate across the treatment groups
- Second time: We balance the first 2 moments (which balances the mean and variance) of each covariate across the treatment groups

```{r}
weight_out_ebal_mom1 <- weightit(
    treat ~ age + educ + race + married + nodegree + re74 + re75,
    data = lalonde,
    estimand = "ATT",
    moments = 1,
    method = "ebal"
)

weight_out_ebal_mom2 <- weightit(
    treat ~ age + educ + race + married + nodegree + re74 + re75,
    data = lalonde,
    estimand = "ATT",
    moments = 2,
    method = "ebal"
)
```

Compare the quality of the weights and the balance statistics across these two implementations:

- In the balance table, how does `Diff.Adj` compare? Does this make sense?
- In the balance table, how does `V.Ratio.Adj` compare? Does this make sense?
- How does the variance of the weights compare between the two approaches?

```{r}
summary(weight_out_ebal_mom1)
bal.tab(
    weight_out_ebal_mom1,
    stats = c("mean.diffs", "variance.ratios")
)

summary(weight_out_ebal_mom2)
bal.tab(
    weight_out_ebal_mom2,
    stats = c("mean.diffs", "variance.ratios")
)
```

